Identifying and correcting multiple sources of misspecification\\in GWAS summary statistics for prediction\\---~OR~---\\Reducing misspecifications and improving tagging\\for better polygenic scores

### 2021-09-02 paper/paper-misspec.tex
finish discu

Therefore, providing an accurate imputation INFO score is important for two reasons. First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
Second, it may lead to an improved prediction when building polygenic scores using one of the proposed corrections.
We apply this correction to the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}). 
All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.

@@

Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced some changes to possibly make these models more robust to these misspecifications.
Previously, we proposed a quality control based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
Here we show that we can refine this quality control and almost get perfect matches by deriving standard deviations from reported allele frequencies in the summary statistics (instead of from the reference panel, which can now be used to compare allele frequencies), and by also correcting for the INFO score (e.g.\ see figure \ref{fig:brca_onco_qc}).
Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics. Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.

Note that, in this study, we use summary statistics for which this information (INFO scores and allele frequencies) is available, but this is far from being the case for all GWAS summary statistics. 
We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score may not be possible; would a weighted averaged INFO score work?
Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
Finally, another information, the effect sample size per variant, is often missing from GWAS summary statistics. 
We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
Again, we note that an overestimated sample size can be detected from the QC plot we propose (when the slope is less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).

[TODO: FINISH]

[TODO: TALK ABOUT HARD INFERRING N != INFO]

[TALK ABOUT SUM(N\_EFF) VS N\_EFF(SUM)?]

[BOTH DISAPOINTING BUT REASSURING]

@@

\begin{table}[h]
	\centering
	\begin{tabular}{|l||c|c|c||c|c|}
		\hline
		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
		\hline
		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,463 & ~~9,874,289 & 1,049,455 & 0.968 \\
		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
		\hline
	\end{tabular}
	\caption{Summary of external GWAS summary statistics used. PRCA summary statistics have many variants with a missing INFO score, which we discard.\label{tab:sumstats}}
\end{table}

>>>>

Therefore, providing an accurate imputation INFO score is important for two reasons. First, it allows for correcting for a reduced standard deviation when using imputed data in the QC step we propose, in order to better uncover problems with the summary statistics.
Second, using one of the proposed corrections may lead to an improved prediction when deriving polygenic scores.
We apply this correction to the two breast cancer summary statistics. We first compare the standard QC proposed in \cite{prive2020ldpred2} (which ends up filtering on MAF here, which we call ``qc1''). We then also filter out the two long-range LD regions of chromosome 6 and 8 for the OncoArray summary statistics and remove around 500 variants when filtering on differences of MAFs between summary statistics and the validation dataset (``qc2'').
As for the correction for the INFO scores, we use the first correction, ``sqrt\_info'', which is simple because it is equivalent to post-processing PGS effects by multiplying them by $\sqrt{\text{INFO}}$.
Although results for both QC used are very similar, correcting for the INFO score slightly improves prediction when deriving polygenic scores based on iCOGS summary statistics (Figure \ref{fig:res_brca}). 
All other improvements introduced before have little or no effect here, probably because misspecifications are much smaller than in the simulations.

@@

Here we have investigated misspecifications in GWAS summary statistics, including when they are derived from imputed data, and when the information on sample size per variant is different for all variants because of meta-analysis.
Previously, we proposed a quality control (QC) based on comparing standard deviations derived from GWAS summary statistics with the ones from a reference panel \cite[]{prive2020ldpred2}.
Here we show that we can refine this quality control by deriving the second standard deviations directly from the reported allele frequencies in the GWAS summary statistics, and by correcting for the INFO scores of the GWAS (e.g.\ see figure \ref{fig:brca_onco_qc}).
Using this refined QC, we were able to identify a potential issue with how principal components were derived in a set of breast cancer summary statistics. Fortunately this has practically no effect on the predictive performance of the derived polygenic scores.
Additional QC can also be performed, e.g.\ comparing reported GWAS allele frequencies with the ones from the LD reference panel, e.g.\ to detect genotyping or allelic errors.
We do perform this additional QC as part of ``qc2'' here, and also when designing the large set of variants from the UK Biobank (Methods). 
One can also run other QC tools such as DENTIST \cite[]{chen2021improved}.

Note that, in this study, we use summary statistics that include extended information (e.g.\ INFO scores and allele frequencies), yet most GWAS summary statistics do not. 
We acknowledge that, in the case of a meta-analysis from multiple studies, providing a single INFO score per variant may not be possible; would a weighted averaged INFO score work?
Nevertheless, this quality control could be easily performed within each study before pooling results, to make sure that summary statistics have the best possible quality for follow-up analyses such as deriving polygenic scores.
Another information, the effective sample size per variant, is often missing from GWAS summary statistics. Even the total effective sample size is sometimes difficult to assess. Indeed, often only the total number of cases and controls are reported in the abstract, and where numbers possibly includes sizes before QC; therefore one has to dig into supplementary materials to find this information.
We recall that if some studies of a meta-analysis have an imbalanced number of cases and controls, the global effective sample size should not be computed from the total numbers of cases and controls overall, but instead from the sum of each effective sample size.
Indeed, take the extreme example of meta-analyzing two studies, one with 1000 cases and 0 controls, and another one with 0 cases and 1000 controls, then the effective sample size of the meta-analysis is 0, not 2000.
Fortunately, an overestimated sample size can be detected from the QC plot we propose (where the slope is then less than 1) for case-control studies using logistic regression; otherwise the standard deviation of the phenotype is also needed (Equation \ref{eq:approx-sd-lin}).

We have assessed the impact of these misspecifications in GWAS summary statistics on the predictive performance of some polygenic score methods.
Using LDpred2 models \cite[]{prive2020ldpred2} and our reimplementation of the lassosum model \cite[]{mak2017polygenic} for deriving polygenic scores, we have introduced and investigated some changes to possibly make these models more robust to misspecifications.
Overall, these changes have provided large improvements of predictive performance in simulations with large misspecifications. However, they have almost no effect when using the published GWAS summary statistics, except for breast cancer and coronary artery disease summary statistics.
This is a little disappointing, but more importantly it is also reassuring because it means that the summary statistics we use in this study are usually of good quality for follow-up analyses such as deriving polygenic scores.

@@

\begin{table}[h]
	\centering
	\begin{tabular}{|l||c|c|c||c|c|}
		\hline
		Trait & GWAS citation & \begin{tabular}{@{}c@{}}Effective GWAS\\sample size\end{tabular} & \# GWAS variants & \begin{tabular}{@{}c@{}}\# matched variants\\with INFO > 0.4\end{tabular} & Mean INFO \\
		\hline
		Breast cancer (BRCA) [iCOGS] & \cite{michailidou2017association} & ~~87,037 & 11,792,542 & 1,051,242 & 0.841 \\
		Breast cancer (BRCA) [OncoArray] & \cite{michailidou2017association} & 104,442 & 11,792,542 & 1,054,233 & 0.968 \\
		Type 1 diabetes (T1D) [Affymetrix] & \cite{censin2017childhood} & ~~~~~5516  & ~~8,996,866 & ~~~934,712 & 0.885 \\	
		Type 1 diabetes (T1D) [Illumina] & \cite{censin2017childhood} & ~~~~~7982  & ~~8,996,866 & ~~~949,334 & 0.942 \\
		Prostate cancer (PRCA) & \cite{schumacher2018association} & 135,316 & 20,370,946 & ~~~818,400 & 0.969 \\
		Depression (MDD) [without UKBB] & \cite{wray2018genome} & 110,464 & ~~9,874,289 & 1,049,455 & 0.968 \\
		Coronary artery disease (CAD) & \cite{nikpay2015comprehensive} & 129,014 & ~~9,455,778 & 1,052,200 & 0.982 \\
		\hline
	\end{tabular}
	\caption{Summary of external GWAS summary statistics used. PRCA summary statistics have many variants with a missing INFO score, which we discard.\label{tab:sumstats}}
\end{table}
